---
title: "Blueprint_isoDu"
author: "Hillary Heiling"
date: "March 21, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CT1 vs CT2

Record results in data.frame

```{r}
filesA = list.files(path = "C:/Users/hheiling/Documents/Longleaf/Blueprint/isoDu_out/CT1_vs_CT2",
                    full.names = T)
length(filesA)

dfA = read.table(filesA[1], header = T)
for(i in 2:length(filesA)){
  df_i = read.table(filesA[i], header = T)
  dfA = rbind(dfA, df_i)
}

dim(dfA)

dfA[1:3,]

# sum(is.na(dfA$pResample))
sum(dfA$pResample < 0.05, na.rm = T)
```

## CT1 vs CT3

Record results in data.frame

```{r}
filesB = list.files(path = "C:/Users/hheiling/Documents/Longleaf/Blueprint/isoDu_out/CT1_vs_CT3",
                    full.names = T)
length(filesB)

dfB = read.table(filesB[1], header = T)
for(i in 2:length(filesB)){
  df_i = read.table(filesB[i], header = T)
  dfB = rbind(dfB, df_i)
}

dim(dfB)

dfB[1:3,]

# sum(is.na(dfA$pResample))
sum(dfB$pResample < 0.05, na.rm = T)
```

## CT2 vs CT3

Record results in data.frame

```{r}
filesC = list.files(path = "C:/Users/hheiling/Documents/Longleaf/Blueprint/isoDu_out/CT1_vs_CT3",
                    full.names = T)
length(filesC)

dfC = read.table(filesC[1], header = T)
for(i in 2:length(filesC)){
  df_i = read.table(filesC[i], header = T)
  dfC = rbind(dfC, df_i)
}

dim(dfC)

dfC[1:3,]

# sum(is.na(dfA$pResample))
sum(dfC$pResample < 0.05, na.rm = T)
```

## Discriminatory Clusters

Clusters that discriminate CT1 from both CT2 and CT3

```{r}
clustsCT1 = intersect(dfA$cluster[which(dfA$pResample < 0.05)],
                      dfB$cluster[which(dfB$pResample < 0.05)])
length(clustsCT1)
```

Clusters that discriminate CT2 from both CT1 and CT3

```{r}
clustsCT2 = intersect(dfA$cluster[which(dfA$pResample < 0.05)],
                      dfC$cluster[which(dfC$pResample < 0.05)])
length(clustsCT2)
```

Clusters that discriminate CT3 from both CT1 and CT2

```{r}
clustsCT3 = intersect(dfB$cluster[which(dfB$pResample < 0.05)],
                      dfC$cluster[which(dfC$pResample < 0.05)])
length(clustsCT3)
```

Clusters that discriminate between all three cell types

```{r}
discrim_clusts = intersect(clustsCT1, intersect(clustsCT2, clustsCT3))
length(discrim_clusts)
```

Since there are well over 100 clusters that are discriminatory between the three cell types, we can try and isolate the clusters with the greatest effect size.

```{r}
dfAA = dfA[which(dfA$cluster %in% discrim_clusts),]
dfBB = dfB[which(dfB$cluster %in% discrim_clusts),]
dfCC = dfC[which(dfC$cluster %in% discrim_clusts),]

p_vals = cbind(dfAA$pResample[order(dfAA$cluster)],
               dfBB$pResample[order(dfBB$cluster)],
               dfCC$pResample[order(dfCC$cluster)])
rownames(p_vals) = discrim_clusts[order(discrim_clusts)]
colnames(p_vals) = c("CT1_CT2","CT1_CT3","CT2_CT3")

ES = pmax(p_vals[,1],p_vals[,2],p_vals[,3])

ES_order = ES[order(ES)]
# minimum of the maximums
min_max = min(ES_order)
sum(ES_order <= min_max)

# Since no way to distinguish between these 300+ clusters, for now record all of these 300+ clusters
discrim_clusts2 = names(ES_order[which(ES_order <= min_max)])

```


Save disciminatory cluster and gene information

```{r}
# load nTE object
load("Materials/gencode.v15.nTE.RData")

nTE_discrim = nTE[which(nTE$clustID %in% discrim_clusts2),]
dim(nTE_discrim)

# Dimension explanation: some clusters have multiple genes
length(unique(nTE_discrim$clustID))
hist(table(nTE_discrim$clustID))

save(nTE_discrim, file = "Materials/gencode.v15.nTE.discriminatory.RData")
```


The End
